Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-06-03 MS/ms.Rmd
Bunch more writing. Sorted out the survey methods description, added to the modelling section

---
title: "Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology"
author: Daniel W.A. Noble, Patrice Potter, Sammy Burke, Szymek Drobniak, Malgorzata Lagisz, Rose O’Dea, Shinichi Nakagawa
date: "`r Sys.Date()`"
bibliography: ../bib/refs.bib
csl: ../bib/the-journal-of-experimental-biology.csl
output: 
  bookdown::word_document2:
    toc: no
    toc_depth: 6
    number_sections: false
    reference_docx: template.docx
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
    bibliography: refs.bib
editor_options: 
  chunk_output_type: console
---

@@

@@

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis, and is becoming increasingly more prominant the field of comparative physiology. Formal meta-analyses involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. The entire process should follow and report upon this entire process carefully to ensure the meta-analysis is as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMAEcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of the research question, along with identifying any gaps and biases, there are two major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analaysis is its ability to account for sampling error []. The inverse sampling error is used to weight higher powered studies in the analysis and make overall esatimtes more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on esatimytes, even if the overall mean itself is unbiased.

@@

## **Survey of Meta-Analytic Practices in Comparative Physiology**

#### *Literature Search and Survey Data Collection*
To understand meta-analytic practices in comparative physiology, we searched for meta-analyses tackling comparative physiological questions that relied on data compiled from studies using animal model systems with more than a single species. All searches were conducted......**write details here**.

We included quantitative syntheses, including meta-analyses (*sensu stricto*), that were focused on comparative physiologcial questions. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data from the literature including traits, such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies, ect.). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival / mortality across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantify plasticity in physiological traits across taxa for the purpose of understanding adaptive plasticity in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source for quantitative synthesis. Quantitative syntheses interested in phenology, biomechanics, sexual size dimorphism (i.e., SSD, except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription and quantitative genetics were excluded. There were also meta-analyses that focused on colour evolution and contrast which varied in their goals but were most often not focused on physiological questions. As such, these studies were also exclude.  We also excluded plant synthesis papers because they were often focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output and greenhouse gas emissions) that did not focus on organism physiological function across species.

@@

## **Opening Up New Opportunities in Comparative Physiology: Expanding the Breadth of Effect Sizes to Deal with Treatment Differences Across Studies**

#### *Brief Introduction to Common Effect Sizes and Sampling Errors that are Useful in Comparative Physiology*
#### *Expanding the Scope of Effect Size Metrics in Comparative Physiology*

@@

@Nakagawa2015 recently proposed analogous effect size estimates to *lnRR* that allow for comparisons of changes in variance between two groups, the log variance ratio (*lnVR*) and the log coefficient of variation (*lnCVR*). Like *lnRR*, *lnVR* and *lnCVR* are ratios that describe the relative difference in trait variablity between two groups. We refer readers to @Nakagawa2015 for the equations describing *lnVR* and *lnCVR*, but these can easily be extended to their $Q_{10}$ analogues (and associated sampling varaince) as follows:

@@

>>>>

---
title: "Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology"
bibliography: ../bib/refs.bib
csl: ../bib/the-journal-of-experimental-biology.csl
output: 
  bookdown::word_document2:
    toc: no
    toc_depth: 6
    number_sections: false
    reference_docx: template.docx
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
    bibliography: refs.bib
editor_options: 
  chunk_output_type: console
---

@@

Daniel W.A. Noble$^{1,*}$, Patrice Potter$^{2}$, Sammy Burke$^{2}$, Szymek Drobniak$^{2}$, Malgorzata Lagisz$^{2}$, Rose O’Dea$^{2}$, Shinichi Nakagawa$^{2}$

$^{1}$ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600

$^{2}$ School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW

* Corespondance:

@@

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominant the field of comparative physiology. Formal meta-analyses involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. The entire process should follow and report upon this entire process carefully to ensure the meta-analysis is as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMAEcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of the research question, along with identifying any gaps and biases, there are two major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analaysis is its ability to account for sampling error [@Koricheva2013]. The inverse sampling error is used to weight higher powered studies in the analysis and make overall esatimtes more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself is unbiased.

@@

## **Survey of Meta-Analytic Practices in Comparative Physiology**
#### *Literature Search and Survey Data Collection*
To understand meta-analytic practices in comparative physiology, we searched for meta-analyses and comparative analyses that tackled questions that relied on using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptable to variation in how treatments are applied across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored and if so it was usable (i.e., in a format like .txt).  All searches were conducted in Scopus between May 16 - 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macrophysiology' in their title and abstract, restricting our search to key comparative physiology journals (deatiled in Table X - Supplementary Materials) within the last 6 years. We expected this to provide a contemporay overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We awknowledge that this may have missed some important, and influential, quantitative synthese in the field of comparative physiology, but likely provided a random sample to inform on the questions we sought to address (see below).

We included quantitative syntheses, including meta-analyses (*sensu stricto*), that were focused on comparative physiologcial questions only. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data from the literature including traits, such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies, ect.). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival / mortality across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantifed plasticity in physiological traits across taxa for the purpose of understanding adaptive plasticity in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source for quantitative synthesis. Quantitative syntheses interested in phenology, biomechanics, sexual size dimorphism (i.e., SSD, except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription and quantitative genetics were excluded. There were also meta-analyses that focused on colour evolution and contrast which varied in their goals but were most often not focused on physiological questions. As such, these studies were also exclude.  We restricted our focus to animal studies because plant synthesis papers were often only focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output and greenhouse gas emissions).

All titles and abstracts and full text screening was done by two people (DWAN and ML). Any conflicts or disagreements were resolved by discussion and a concensus on a paper was reached. Out of the 417 papers, 79 full texts were screened and we included a total of 62 for final data extraction. For full details on the search strings used, dates searched, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the *Supplementary Materials*.

@@

## **Opening Up New Opportunities in Comparative Physiology: Expanding the Breadth of Effect Sizes to Deal with Treatment Differences Across Studies**
#### *Brief Introduction to Common Effect Sizes and Sampling Errors that are Useful in Comparative Physiology*

#### *Expanding the Scope of Effect Size Metrics in Comparative Physiology*

@@

@Nakagawa2015 recently proposed analogous effect size estimates to *lnRR* that allow for comparisons of changes in variance between two groups, the log variance ratio (*lnVR*) and the log coefficient of variation (*lnCVR*). Like *lnRR*, *lnVR* and *lnCVR* are ratios that describe the relative difference in trait variablity between two groups. We refer readers to @Nakagawa2015 for the equations describing *lnVR* and *lnCVR*, but these can easily be extended to their $Q_{10}$ analogues (and associated sampling variances) as follows:

@@

If we assume that we are interested in understanding the overall effect of, say, stress (i.e., corticosterone) on XX we might extract data from studies manipulating corticosterone on eggs. Of course, these studies are diverse, apply different dosages of corticosterone and are working across many species. If the studies are experimental in nature, than a standardised effect size, such as $ln_{RR}$ might be applicable. We then might fit an overall, multilevel meta-analytic model as follows:

$$ z_{i} = u + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i} $$

where $z_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), *u* the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect for species *k* applied to effect size *i*. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_aA$ [hereafter denoted ~ *N*$(0, \sigma^{2}_aA)$], where A is a phylogenetic correlation matrix derived from a phylogenetic tree, $sp_{k[i]}$ is the species-specific effect for the *k*th species applied to effect size *i* [assumed to be ~ *N*$(0, \sigma^{2}_{sp}I)$], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i* [assumed to be ~ *N*$(0, \sigma^{2}_{s}I)$, where I is an identity matrix], $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size [assumed to be ~ *N*$(0, \sigma^{2}_eE)$], and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect size [assumed to be  ~ *N*$(0, M)$], where M is a variance-covariance matrix with the diagonal elements being the effect size sampling variance. This model is a 'weighted' multilevel meta-analysis model because we can estimate the overall sampling variance. This is very important because it provides a way for us to understand just how much variation in effects are simplly the result of sampling differences across studies versus other, potentially improtant sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe. This is formalised in what we call 'heterogenetiy analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating corticosterone, how much of an overall effect does corticostrone have on XX relative to control conditions?
